---
date: "2020-04-22"
diagram: true
# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image: 
  caption: ''
  focal_point: Smart
  preview_only: true
title: 'Scikit Learn in R'
author: "Jim Gruman"
output: 
 blogdown::html_page:
  toc: false
categories: [Python, Data Science, R]
description: "Setup to Run Python in RStudio"
featured: false
draft: false
---

>Inspired by [Matt Dancho's Business-Science Learning Series post](https://www.business-science.io/learn-r/2020/04/20/setup-python-in-r-with-rmarkdown.html)

The 2 most popular data science languages - `Python` and `R` - are often characterized as rivals. This couldn’t be further from the truth. Data scientists use the strengths of each language environment. For Example:

- **Machine Learning**: They can switch to Python to leverage Scikit Learn and PyTorch.

- **Data Wrangling, Visualization, Apps & Reporting**: They can quickly change to R to use `tidyverse`, `shiny` and `rmarkdown`.

The bottom line is that knowing both R and Python makes teams much more productive. 

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

This post is a quick demonstration of the setup and creation of an environment to use Python from within the RStudio IDE using the `reticulate` package.

Step 1: Install the Anaconda distribution

Step 2: Install the `reticulate` R package

```{r}
library(tidyverse)
library(reticulate)
```

As of today, Anaconda still ships with python 3.7.

To update to python 3.8, run the following code at the anaconda prompt:

>`conda install -c anaconda python=3.8`

>`conda update --all`

>`conda create -n py3.8 python=3.8 scikit-learn pandas numpy matplotlib`

>`conda activate py3.8`

>`python --version`

List your Conda Environments, from RStudio:

```{r}
reticulate::conda_list()

```

Set Your Conda Environment, and double check that reticulate is actually using the new conda env.

```{r}
reticulate::use_condaenv("py3.8", required = TRUE)

py_config()

# Another setup approach -
#reticulate::conda_install(packages = c("scikit-learn", "pandas", "numpy", "matplotlib"))
#.
#at installation, watch for a message to launch an Anaconda Prompt to update conda by typing "conda update -n base -c defaults conda"
```

## Python tests

Is python working?

```{python}
1 + 1

```


Sourcing an existing python script is as easy as calling `reticulate::source_python()`. For example, if you have the following Python script *flights.py*

```{python eval=FALSE}
import pandas
def read_flights(file):
  flights = pandas.read_csv(file)
  flights = flights[flights['dest'] == "ORD"]
  flights = flights[['carrier', 'dep_delay', 'arr_delay']]
  flights = flights.dropna()
  return flights
```

Then you can source the script and call the `read_flights()` function as follows:

```{r eval=FALSE}
source_python("flights.py")
flights <- read_flights("flights.csv")

library(ggplot2)
ggplot(flights, aes(carrier, arr_delay)) + geom_point() + geom_jitter()
```

Import numpy and pandas using the import shorthand `np` and `pd` respectively. 

- numpy: Math Calculations
- pandas: Data Wrangling

```{python}
import numpy as np
import pandas as pd
import os
os.environ['QT_QPA_PLATFORM_PLUGIN_PATH'] = 'C:\\Users\\jimgr\\Anaconda3\\Library\\plugins\\platforms'
```

### Testing numpy

```{python}
np.arange(1, 10)
```

### Testing pandas

```{python}
# Make a sequence in a data frame using dict format
df = pd.DataFrame(data = {"sequence":np.arange(1,20,.01)})

# Use assign (mutate) equivalent to calculate the np.sin() of the series
df = df.assign(value=np.sin(df["sequence"]))

df
```

### Test Matplotlib

```{python}
import matplotlib.pyplot as plt

t = np.arange(0.0, 2.0, 0.01)
s = 1 + np.sin(2*np.pi*t)
plt.plot(t, s)
plt.xlabel('time(s)')
plt.ylabel('voltage(mV)')
plt.grid(True)
plt.savefig("Matplotlibtest.png")
plt.show()
```

### Test Scikit Learn

```{python}
from sklearn.ensemble import RandomForestClassifier

clf = RandomForestClassifier(random_state=0)

X = [[ 1,  2,  3],  # 2 samples, 3 features
     [11, 12, 13]]

y = [0, 1]  # classes of each sample

clf.fit(X, y)

clf.predict(X)  # predict classes of the training data
```


```{python}
from sklearn.cluster import AffinityPropagation
from sklearn.datasets import make_blobs

# Generate sample data
centers = [[1, 1], [-1, -1], [1, -1]]
X, labels_true = make_blobs(n_samples=300, centers=centers, cluster_std=0.5,
                            random_state=0)

# Compute Affinity Propagation
af = AffinityPropagation(preference=-50).fit(X)
cluster_centers_indices = af.cluster_centers_indices_
labels = af.labels_

n_clusters_ = len(cluster_centers_indices)

# Plot result
import matplotlib.pyplot as plt
from itertools import cycle

plt.close('all')
plt.figure(1)
plt.clf()

colors = cycle('bgrcmykbgrcmykbgrcmykbgrcmyk')
for k, col in zip(range(n_clusters_), colors):
    class_members = labels == k
    cluster_center = X[cluster_centers_indices[k]]
    plt.plot(X[class_members, 0], X[class_members, 1], col + '.')
    plt.plot(cluster_center[0], cluster_center[1], 'o', markerfacecolor=col,
             markeredgecolor='k', markersize=14)
    for x in X[class_members]:
        plt.plot([cluster_center[0], x[0]], [cluster_center[1], x[1]], col)

plt.title('Estimated number of clusters: %d' % n_clusters_)
plt.show()
```

----

### Did you find this page helpful? Consider sharing it 🙌



