---
title: "rtweet"
author: "Jim Gruman"
date: "`r Sys.Date()`"
diagram: true
output: 
 blogdown::html_page:
  toc: false
categories: [R]
description: "Analyze Twitter data" 
image: 
  caption: ''
  focal_point: "Smart"
  preview_only: true
featured: false
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)
```

I was inspired by David Robinson's Tidy Tuesday screencast to work with `rtweet` on some familiar Twitter handles and ended up discovering a bug and a workaround for windows users. `rtweet` is an amazing way of having a little productive procrastination while honing R skills. 

In this post, we will look at the accounts that I engage most often, the time of day for my Twitter activity, and build a `reactable` interactive table to sort and filter on the hashtag #TheDayTheFarmChanged. 

----



As with any other project, start with loading the appropriate libraries.

```{r, echo=TRUE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)   # tidy data manipulation, including dplyr and ggplot
library(rtweet)      # the R Twitter API
theme_set(hrbrthemes::theme_ft_rc())   # Set a ggplot color, font theme
library(showtext)   
library(lubridate)   # date manipulation tools
showtext_auto()
library(reactable)   # A Formatted Sortable Interactive Table 
```

The `rtweet` [Obtaining and using access tokens](http://rtweet.info/articles/auth.html) vignette gives instructions for creating tokens and writing them to `.Renviron`, one-time only, as:

```{r eval=FALSE}
token <- create_token(
  app = "mytwitterapp_NDMSBA",
  consumer_key = api_key,
  consumer_secret = api_secret_key,
  access_token = access_token,
  access_secret = access_token_secret)
```
Unfortunately, the process currently works properly in Linux environments, but not in Windows. Closely inspect the token created:

```{r eval=FALSE}
 token<-rtweet::get_token()

 token
```

You have found the bug when the token returned does not match the keys created previously.

The solution to the problem is to edit your local `.Renviron` file manually. In the console, type

```{r eval=FALSE}
usethis::edit_r_environ()
```

Find the **TWITTER_PAT** parameter and change every back slash `\` and forward slash `/` to double backslashes `\\` to the rds file address.  It will resemble

`"C:\\Users\\<youruser>\\Documents\\.rtweet_token.rds"`

Save the `.Renviron` file.  Then Restart R (Ctrl+Shift+F10). Closely inspect the token again:

```{r eval=FALSE}
token<-rtweet::get_token()

token
```
It turns out that [Jon Harmon has also reported of the issue at the package Github repository.](https://github.com/ropensci/rtweet/issues/380). 

----

Let's have a look at the accounts that I favorite (or **like**) the most:

```{r }
# there is a rate limit at Twitter at approximately 18000 each 15 minutes
df_likes <- rtweet::get_favorites("jim_gruman", n = 3000)
```

```{r warning=FALSE, fig.align='center', fig.width=7, fig.height=5}
users_favorite<-df_likes %>% 
    group_by(screen_name, user_id) %>%
    summarise(
        quant_fav= n()
    )

users_favorite %>%
    ungroup() %>%
    top_n(15,quant_fav) %>%
    mutate(screen_name = reorder(screen_name,quant_fav)) %>%
    ggplot(aes(y = screen_name, x = quant_fav, fill = quant_fav)) + 
    geom_col(show.legend = FALSE) +
    theme(plot.title.position = "plot",
          text = element_text(hrbrthemes::font_rc, size = 16))+
    labs(
        title = "Top 15 Twitter accounts I like the most often",
        subtitle = paste0("Within the past 3000 Tweets, since ",min(as_date(df_likes$created_at,  tz ="America/Chicago"))),
        y = "@ user names",
        x= "# of tweets liked",
        caption = paste0("@jim_gruman | ", Sys.Date())
    )

```

And let's break down my activity by week day

```{r warning=FALSE, fig.align='center', fig.width=7, fig.height=5}
tw_activity<- df_likes %>% 
    transmute(
              tw_hour = hour(as_datetime(created_at, tz ="America/Chicago")),
              tw_day = as_date(created_at, tz ="America/Chicago"),
              tw_weekday = wday(tw_day,label = T, abbr = T, week_start = 1)
    ) %>%
    group_by(tw_weekday, tw_hour) %>%
    summarise(
        quant_fav= n()
    ) %>% 
    ungroup() %>% 
    group_by(tw_weekday) %>% 
    mutate(prop = quant_fav / sum(quant_fav)) %>% 
    ungroup()
  
tw_activity %>%
    ggplot(aes(x = tw_hour, y = prop, fill = prop)) + 
    geom_col(position = position_nudge(x = .5),
             show.legend = FALSE) +
    scale_x_continuous(breaks = c(6, 12, 18))+
    hrbrthemes::scale_y_percent()+
    facet_wrap(~tw_weekday, ncol = 5)+
    theme(text = element_text(hrbrthemes::font_rc, size = 16),
    panel.grid.minor = element_blank(),
    plot.title.position = "plot",
    plot.margin = margin(0.25,0.5,0.25,0.5,"in")
      )+
    labs(
        title = "My recent liking activity on Twitter",
        subtitle = paste0("Within the past 3000 Tweets, since ",min(as.Date(df_likes$created_at, tz = "America/Chicago"))),
        x = "hour of the day",
        y = "",
        caption = paste0("@jim_gruman | ", Sys.Date())
    )

```

Some may prefer the `ggridges` version of the same data, visualized a bit differently.

```{r warning=FALSE, fig.align='center', fig.width=7, fig.height=5}
library(ggridges)

tw_activity %>%
  mutate(tw_weekday = fct_rev(tw_weekday))%>%
ggplot(aes(x = tw_hour, , y = tw_weekday, height = prop, fill = stat(x)))+
ggridges::geom_ridgeline_gradient(scale = 10, show.legend = FALSE)+
  scale_fill_gradient2(low = "blue",
                        mid = "grey50",
                        high = "blue",
                        midpoint = 12L)+
    theme(text = element_text(hrbrthemes::font_rc, size = 16),
    panel.grid.minor = element_blank(),
    plot.title.position = "plot",
    plot.margin = margin(0.25,0.5,0.25,0.5,"in")
      )+
    labs(
        title = "My recent liking activity on Twitter",
        subtitle = paste0("Within the past 3000 Tweets, since ",min(as.Date(df_likes$created_at, tz = "America/Chicago"))),
        x = "hour of the day",
        y = "Proportion of Tweet Likes",
        caption = paste0("@jim_gruman | ", Sys.Date())
    )

```

----

Finally, let's build a `reactable` table to sort and search for individual tweets for the #TheDayTheFarmChanged hastag.

```{r}
# the URL column returns a list that has to be unlisted. This is a helper function
make_url_html <- function(url) {
  if(length(url) >2) {
    if(!is.na(url)) {
      as.character(glue::glue("<a title = {url} target = '_new' href = '{url}'>{url}</a>") )
    } else {
        ""
      }
  } else {
    paste0(purrr::map_chr(url, ~ paste0("<a title = '", .x, "' target = '_new' href = '", .x, "</a>", collapse = ", ")), collapse = ", ")
  }
}

## Search returns tweets from the past 6-9 days

tweet_search_data <-search_tweets("#TheDayTheFarmChanged", n = 200, include_rts = FALSE)  %>%
  select(user_id, status_id, created_at, screen_name, text, favorite_count, retweet_count, urls_expanded_url) %>%
  mutate(
    Tweet = glue::glue("{text} <a href='https://twitter.com/{screen_name}/status/{status_id}'>>> </a>") 
    )%>%
  select(DateTime = created_at, User = screen_name, Tweet, Likes = favorite_count, RTs = retweet_count, URLs = urls_expanded_url)

# Convert the URLs from the list column into HTML links
tweet_search_data$URLs <- purrr::map_chr(tweet_search_data$URLs, make_url_html)

reactable(tweet_search_data,
            filterable = TRUE,
            searchable = TRUE,
            bordered = TRUE,
            striped = TRUE,
            highlight = TRUE,
            defaultPageSize = 25, 
            showPageSizeOptions = TRUE,
            showSortable = TRUE,
            pageSizeOptions = c(25,50,100),
            defaultSortOrder = "desc",
            columns = list(
              DateTime = colDef(defaultSortOrder = "asc"),
              User = colDef(defaultSortOrder = "asc"),
              Tweet = colDef(html = TRUE, minWidth = 190, resizable = TRUE),
              Likes = colDef(filterable = FALSE),
              RTs = colDef(filterable = FALSE),
              URLs = colDef(html = TRUE)
            )
)

```

----

### Did you find this page helpful? Consider sharing it 🙌

