---
title: 'R 4.0.0 Arbor Day'
author: "Jim Gruman"
date: "2020-04-23"
diagram: true
output: 
 blogdown::html_page:
  toc: false
categories: [R]
description: "Significant User-Visible Changes are Coming" 
# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image: 
  caption: ''
  focal_point: "TopRight"
  preview_only: false
featured: false
draft: false
---

R 4.0.0 is expected to be released to CRAN on Friday April 24th

At that time, all packages will need to be (re-)installed.

Here's a short script to keep a list of your current R packages and how to reinstall them after updating:

```{r eval=FALSE}
# Before the update, save current packages and their versions to ip
ip <- installed.packages()
ip

saveRDS(ip, "~/CurrentPackages.rds")
```

```{r eval=FALSE}
# After updating base R, load the file in reverse

ip <- readRDS("~/CurrentPackages.rds")
 
install.packages(ip[,1])

```

[The list of R 4.0.0 Features, Utilities, and Bug Fixes](https://stat.ethz.ch/R-manual/R-devel/doc/html/NEWS.html)

On my windows system, I am going to start fresh. It will be interesting to come back in a month to compare what I need with what I had previously.

----

### Did you find this page helpful? Consider sharing it 🙌


