---
date: "2022-06-16"
title: 'Generations in America'
author: "Jim Gruman"
format: hugo
categories: [R]
summary: "The composition of American generations per Pew Research & US Census"
featured: false
draft: false
---

Inspired and updated from

> [Jason Timm's post](https://www.jtimm.net/2020/06/10/american-generations/)

> Kyle Walker's [Analyzing US Census Data](https://walker-data.com/census-r/index.html) updates 

### American Generations

Let's have a look at the composition of American generations with Pew Research definitions and US Census data. 

The outcome: a perspective on the changing demographics of America.

Four of America’s seven living generations are more or less “complete,” and only getting smaller: `Greatest`, `Silent`, `Boomers`, and `Gen X`. The generation comprised of `Millenials` is complete as well, in that it has been delineated chronologically; however, the group continues to grow via immigration.

```{r }
#| label: setup
#| include: false
#| code-fold: true
#| code-summary: "Show code"

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r}
#| label: load
#| code-fold: true
#| code-summary: "Show code"

suppressPackageStartupMessages({
library(tidyverse) # tidy data manipulation, including dplyr and ggplot
library(lubridate)   # date manipulation tools
library(tidycensus)
  
library(tigris)

})
theme_set(hrbrthemes::theme_ipsum_tw())   # Set a ggplot color, font theme
showtext::showtext_auto() # helps to locate windows fonts

options(tigris_use_cache = TRUE, 
        tigris_class = "sf")

gen <- c('Post-Z',
         'Gen Z',
         'Millennial',
         'Gen X',
         'Boomers',
         'Silent',
         'Greatest')

range <- c('> 2012',    # post Z
           '1997-2012', # Gen Z
           '1981-1996', # Millenials
           '1965-1980', # X
           '1946-1964', # Boomers
           '1928-1945', # Silent
           '< 1927')    # greatest

gen_desc <- tibble(rank = 7:1,
                   gen = gen,
                   range = range) %>%
  arrange(rank)

gen_desc %>%
  knitr::kable(format = "html",
               booktabs = T,
               caption = "Pew Research Definitions") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F
  ) %>%
  kableExtra::add_footnote(
    c(
      "https://www.pewresearch.org/fact-tank/2019/01/17/where-millennials-end-and-generation-z-begins/"
    )
  )

```

While `Gen Z` has been tentatively stamped chronologically by the folks at Pew Research, only about half have entered the work force. And though we include them here, the `Post-Z` generation is mostly but a twinkle in someones eye; half of the group has yet to be born.

### Monthly US population estimates

The *Monthly Postcensal Resident Population plus Armed Forces Overseas, December 2021* is made available by the US Census [here](https://www.census.gov/programs-surveys/popest/technical-documentation/research/evaluation-estimates/2020-evaluation-estimates/2010s-national-detail.html). The census has transitioned to a new online interface, and this particular file is built off of some of their projections.

```{r}
#| label: readCensus
#| code-fold: true
#| code-summary: "Show code"
pops <- read_csv(
  file = "https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/national/asrh/NC-EST2020-ALLDATA-R-File24.csv",
  show_col_types = FALSE) %>% 
  filter(MONTH == '12' & YEAR == '2021') |> 
  pivot_longer(names_to = "race",
               values_to = "pop",
               cols = c(-UNIVERSE:-AGE)) |> 
  replace_na(replace = list(pop = 0))
```

A more detailed description of the population estimates can be found [here](https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2010-2018/nc-est2018-alldata.pdf). Note: Race categories reflect non-Hispanic populations.

```{r}
#| label: prepForJoin
#| code-fold: true
#| code-summary: "Show code"
race <- c('NHWA', 'NHBA', 'NHIA', 
          'NHAA', 'NHNA', 'NHTOM', 'H')

race1 <- c('White Alone',
           'Black Alone',
           'American Indian Alone',
           'Asian Alone',
           'Native Hawaiian Alone',
           'Two or More Races',
           'Hispanic')

labels <- tibble(race = race, 
                 race1 = race1)

search <- paste(paste0('^',race, '_'), collapse =  '|')
```

The following table details **a random sample of the data set** – with Pew Research defined generations & estimated year-of-birth.

```{r}
#| label: genPops
#| code-fold: true
#| code-summary: "Show code"
gen_pops <- pops %>%
  filter(grepl(search, race)) %>%
  mutate(race = gsub('_.*$', '', race)) %>%
  group_by(AGE, race) %>%
  summarise(pop = sum(pop),
            .groups = "drop") %>%
  left_join(labels, by = "race") %>%
  filter(AGE != '999') %>%
  mutate(yob = 2021 - AGE)  %>% 
  mutate(gen = case_when(
    yob > 2012 ~ 'Post-Z',
    yob < 2013 & yob > 1996 ~ 'Gen Z',
    yob < 1997 & yob > 1980 ~ 'Millennial',
    yob < 1981 & yob > 1964 ~ 'Gen X',
    yob < 1965 & yob > 1945 ~ 'Boomers',
    yob < 1946 & yob > 1927 ~ 'Silent',
    yob < 1928 ~ 'Greatest',
    TRUE ~ "big problem")) %>%
  left_join(gen_desc, by = "gen") %>%
  ungroup() %>%
  select(gen, rank, range, race, 
         race1, yob, AGE, pop)

set.seed(999)
gen_pops %>% sample_n(7)  %>%
  select(gen, range, race1:pop) %>%
  knitr::kable(format = "html",
               booktabs = T,
               caption = "Seven Sample Census Data Lines with Pew Generational Definitions") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F
  ) 
```

### Compositions of American Generations

The figure below summarizes the US population by generation. These numbers will vary some depending on the data source. 

```{r }
#| label: barchart
#| fig.asp: 1
#| code-fold: true
#| code-summary: "Show code"

gen_pops %>%
  group_by(gen, rank) %>%
  summarize(pop = sum(pop),
            .groups = "drop") %>%
  mutate(lab = round(pop/1000000, 1)) %>%
  ggplot(aes(x = pop, 
             y = fct_reorder(gen, rank), 
             fill = gen)) +
  geom_col(show.legend = FALSE)  +
  geom_text(aes(label = lab), 
            size = 6,
            nudge_x = 3000000) +
  hrbrthemes::scale_fill_ipsum() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title.position = "plot",
        axis.text.y = element_text(size = 6*3)) +
  labs(title = 'Population by American Generation in Millions',
       caption = 'SOURCE: US Census, Monthly Postcensal Resident Population plus Armed Forces Overseas, December 2021.',
       x = NULL, 
       y = NULL)
```

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| label: annotationHelper

gg <- gen_pops %>% 
  group_by(yob, AGE, gen) %>%
  summarize(tot = sum(pop),
            .groups = "drop") %>%
  group_by(gen) %>%
  mutate(tot = max(tot)) %>% 
  filter(yob %in% c('1919', '1928', '1946', '1965', 
                    '1981', '1997', '2013'))
```

The figure below illustrates the US population by single year of age, ranging from the population aged less than a year to the population over 100. Generation membership per single year of age is specified by color.

```{r }
#| label: singleyear
#| fig.asp: 1
#| code-fold: true
#| code-summary: "Show code"

gen_pops %>%
  ggplot(aes(x = AGE, 
             y = pop, 
             fill = gen)) +
  geom_vline(xintercept = gg$AGE,
             linetype = 2, 
             color = 'gray', 
             size = .25) +
  geom_col(show.legend = FALSE,
           width = .7)   +
  annotate(geom = "text", 
           x = gg$AGE - 8, 
           y = gg$tot + 100000, 
           label = gg$gen,
           size = 8) +
  hrbrthemes::scale_fill_ipsum() +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text.x = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        plot.title.position = "plot") +
  scale_x_reverse(breaks = rev(gg$AGE)) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(title = 'American Population by Single-year Age & Generation',
       caption = 'SOURCE: US Census, Monthly Postcensal Resident Population plus Armed Forces Overseas, December 2021.',
       x = "Age",
       y = NULL)
```
Next, let's crosscut the single year of age counts presented above by race & ethnicity.

```{r}
#| label: ethnicity
#| fig.asp: 1
#| code-fold: true
#| code-summary: "Show code"

gen_pops %>%
  ggplot(aes(x = AGE, 
             y = pop, 
             fill = race1)) +
  geom_area(stat = "identity",
            color = 'white') +
  geom_vline(xintercept = gg$AGE,
             linetype = 2, color = 'gray', size = .25) +
  annotate(geom = "text", 
           x = gg$AGE - 8, 
           y = gg$tot + 100000, 
           label = gg$gen,
           size = 8) +
  hrbrthemes::scale_fill_ipsum(guide = guide_legend(label.position = "bottom",
             nrow = 1, 
             label.theme = element_text(size = 18) )) +
  theme(legend.position = "bottom",
        legend.justification = "center",
        legend.title = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text.x = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        plot.title.position = "plot",
        legend.key.width = unit(5, "line")) +
  scale_x_reverse(breaks = rev(gg$AGE) ) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(title = 'US Population by Age, Race & Generation',
       caption = 'SOURCE: US Census, Monthly Postcensal Resident Population \nplus Armed Forces Overseas, December 2021.',
       x = "Age",
       y = NULL)

```


```{r}
#| code-fold: true
#| code-summary: "Show code"

white_label <- gen_pops %>% 
  group_by(gen, AGE) %>%
  mutate(per = pop/sum(pop)) %>%
  filter(race1 == 'White Alone') %>%
  group_by(gen) %>%
  mutate(per = max(per)) %>% #For labels below.
  arrange(yob) %>%
  filter(yob %in% c('1919', '1928', '1946', '1965', 
                    '1981', '1997', '2013'))
```

The last figure illustrates a **proportional perspective of race & ethnicity in America** by single year of age. Per figure, generational differences (at a single point in time) can shed light on (the direction of) potential changes in the overall composition of a given populace, as well as a view of what that populace may have looked like in the past.

```{r}
#| label: ethnicityProportion
#| fig.asp: 1
#| code-fold: true
#| code-summary: "Show code"

gen_pops %>%
  group_by(gen, AGE) %>%
  mutate(per = pop/sum(pop)) %>%
  ggplot(aes(x = (AGE), 
             y = per, 
             fill = race1)) +
  geom_area(stat = "identity",
            color = 'white',
            alpha = 0.85) +
  geom_hline(yintercept = .5, 
             linetype = 4,
             color = 'white') +
  geom_vline(xintercept = gg$AGE,
             linetype = 2, 
             color = 'gray', 
             size = .25) +
  annotate(geom = "text", 
           x = gg$AGE - 4.5, 
           y = white_label$per - .09, 
           label = gg$gen,
           size = 8,
           color = "black") +
  hrbrthemes::scale_fill_ipsum(guide = guide_legend(label.position = "bottom",
             nrow = 1,
             label.theme = element_text(size = 18)
             )) +
  theme(legend.position = "bottom",
        legend.justification = "center",
        legend.title = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.text.x = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        plot.title.position = "plot",
        legend.key.width = unit(5, "line")) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_reverse(breaks = rev(gg$AGE)) +
  labs(title = 'Portion of American Population by Age, Race & Generation',
       fill = "",
       x = "Age",
       y = NULL)

```

### American generations in (apparent) time & space

Aggregate race-ethnicity profiles for America’s seven living generations are presented in the table below. Per [American Community Survey (ACS) 2018 5-year estimates](https://github.com/jaytimm/tech-notes-american-politics/blob/master/notes/America_american.md), the `Gen X` race-ethnicity profile (or distribution) is most representative of American demographics overall.

```{r}
#| code-fold: true
#| code-summary: "Show code"

sums <- gen_pops %>%
  group_by(rank, gen, race1) %>%
  summarize(pop = sum(pop),
            .groups = "drop") %>%
  group_by(gen) %>%
  mutate(per = round(pop / sum(pop), 3))

sums %>%
  select(rank, gen, race1, per) %>%
  mutate(per = scales::percent(per, accuracy = 0.1)) %>%
  pivot_wider(names_from = "race1",
              values_from = "per") %>%
  knitr::kable(format = "html",
               booktabs = T,
               caption = "Race-Ethnicity Distribution of American Generations") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F
  )  %>%
  kableExtra::row_spec(4, color = "white", background = "#D18975")
```

### Race-ethnicity profiles for US Counties

Here, we compare race-ethnicity profiles for US counties to those of American generations. Using the `tidycensus` R package, we first obtain county-level race-ethnicity estimates (ACS 2020 5-year);

```{r}
#| code-fold: true
#| code-summary: "Show code"

#x <- tidycensus::load_variables(year = 2018, dataset = 'acs5/profile')
variable <- c('DP05_0071P', 'DP05_0077P', 'DP05_0078P',
              'DP05_0079P', 'DP05_0080P', 
              'DP05_0081P', 'DP05_0082P', 
              'DP05_0083P')

variable_name <- c('Hispanic', 'White Alone', 'Black Alone','American Indian Alone', 'Asian Alone','Native Hawaiian Alone', 'Some Other Race Alone','Two or More Races')

gen <-  tidycensus::get_acs(geography = 'county',
                            variables = variable,
                            year = 2020,
                            survey = 'acs5') %>%
  left_join(tibble(variable, variable_name)) %>%
  select(-variable, -moe) %>%
  rename(variable = variable_name) %>%
  select(GEOID:NAME, variable, estimate)
```

then, via the `tigris` library, we obtain shapefiles for (1) all US counties and (2) US states.

```{r}
#| code-fold: true
#| code-summary: "Show code"

nonx <- c('78', '69', '66', '72', '60', '15', '02')

us_counties <- tigris::counties(cb = TRUE) %>% 
  filter(!STATEFP %in% nonx)
us_states <- tigris::states(cb = TRUE) %>% 
  filter(!STATEFP %in% nonx)

laea <- sf::st_crs("+proj=laea +lat_0=30 +lon_0=-95")
us_counties <- sf::st_transform(us_counties, laea)
us_states <- sf::st_transform(us_states, laea)
```

### Comparing Distributions

The table below highlights the race-ethnicity distribution for my hometown, in **Buchanan County, IA**, GEOID = 19019.

```{r }
#| code-fold: true
#| code-summary: "Show code"

gen %>%
  filter(GEOID == "19019",
         estimate > 0) %>%
  mutate(estimate = scales::percent(estimate / 100, accuracy = 0.1)) %>%
  pivot_wider(names_from = "variable",
              values_from = "estimate") %>%
  select(-GEOID,-NAME) %>%
  knitr::kable(format = "html",
               booktabs = T,
               caption = "Race-Ethnicity Distribution of Buchanan County, Iowa") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F
  ) 

```

Via `Kullback–Leibler divergence` (ie, relative entropy), we compare the demographic profile of Buchanan County to the demographic profiles of each American generation. Per the table below, the Buchanan County profile is most similar to that of the `Greatest` generation. Per the lowest relative entropy value. This is not to say that there are more `Greatest` generation in Buchanan County; instead, the racial demographics of Buchanan County are most akin to an America when the `Greatest` generation were in their prime.

```{r}
#| code-fold: true
#| code-summary: "Show code"

which1 <- gen %>%
  mutate(estimate = ifelse(estimate == 0, .0000001, estimate)) %>%
  inner_join(sums, by = c('variable' = 'race1')) %>%
  
  mutate(div = estimate/100 * log((estimate/100)/per)) %>%
  group_by(GEOID, NAME, gen) %>%
  summarize(relative_entropy = round(sum(div), 4),
            .groups = "drop")

which1 %>%
  filter(GEOID == "19019") %>%
  left_join(gen_desc) %>%
  select(rank, NAME, gen, relative_entropy) %>%
  arrange(rank) %>%
  knitr::kable(format = "html",
               booktabs = T,
               caption = "Kullback–Leibler Divergence of Buchanan County, Iowa") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F
  )  %>%
  kableExtra::row_spec(1, color = "white", background = "#D18975")

```

For reference, the table below highlights the generational distribution for **Buchanan County, IA**, GEOID = 19019.

```{r}
#| code-fold: true
#| code-summary: "Show code"

which1 %>%
  filter(GEOID == "19019") %>%
  mutate(relative_entropy = scales::percent(relative_entropy, accuracy = 0.1)) %>%
  pivot_wider(names_from = "gen",
              values_from = "relative_entropy") %>%
  select(-GEOID,-NAME) %>%
  knitr::kable(format = "html",
               booktabs = T,
               caption = "Generational Distribution of Buchanan County, Iowa",
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F
  ) 
```

```{r}
#| code-fold: true
#| code-summary: "Show code"

which2 <- which1 %>%
  group_by(GEOID, NAME) %>%
  filter(relative_entropy == min(relative_entropy)) %>%
  ungroup() 
```

So, while `Gen X` is most representative of America in the aggregate, per the map below, the demographics of America’s youngest & eldest generations are often most prevalent within each county-level unit. `Post-Z` in the West, and the `Greatest` in the East & Midwest. New & Old Americas, perhaps.

```{r}
#| label: uscounties
#| fig.asp: 1
#| code-fold: true
#| code-summary: "Show code"

us_counties %>%
  left_join(which2, by = 'GEOID') %>%
  ggplot() + 
  geom_sf(aes(fill = gen),  lwd = 0) + 
  geom_sf(data = us_states, 
          fill = NA, 
          show.legend = F, 
          color = "darkgray", 
          lwd = .5) +
  hrbrthemes::scale_fill_ipsum(guide =
  guide_legend(label.position = "bottom", 
             nrow = 1, 
             keywidth = 5,
             label.theme = element_text(size = 18))) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major =  element_blank(),
        legend.position = 'bottom',
        legend.justification = "center",
        plot.title.position = "plot",
        legend.key.width = unit(0.5, "cm")) +
  labs(title = "Counties & their Most Representative Generational Proportions",
       fill = "")
```

### Conclusion

There are many ways to characterize the different perspectives on the composition of America & American generations, and the changing demographics of America.

----

### Did you find this page helpful? Consider sharing it 🙌



